---
title: "DOT: A flexible multi-objective optimization framework for transferring features across single-cell and spatial omics"
author:
  - name: Arezou Rahimi
affiliation: 
  - Heidelberg University Hospital, Heidelberg, Germany
email: arezou.rahimi@uni-heidelberg.de
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{general}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, message = FALSE, #cache = TRUE,
  comment = "#>"
)
```
# Setup

```{r setup, message=FALSE}
# dot
library(DOT)

# for plotting
library(ggplot2)
```

The following example illustrates how DOT is used for inferring cell type composition of spots in a synthetic multicell spatial data of the primary motor cortex region (MOp) of the mouse brain. The reference single-cell RNA seq sample comes from a similar region and contains 44 cell types. Both spatial and single-cell data are down-sampled to 500 genes in this example.


```{r ref, fig.height=5, fig.width=9.5, fig.align='left'}
data(dot.sample)

# gene x cell
dim(dot.sample$ref$counts)

ggplot(as.data.frame(sort(table(dot.sample$ref$labels), decreasing = T)), aes(x = Var1, y = Freq))+
  geom_bar(stat="identity", fill="steelblue")+
  xlab("")+ylab("")+theme_bw()+
  theme(panel.background = element_rect(fill = 'white'), 
        panel.grid = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

```{r srt, fig.height=5, fig.width=6, out.width="70%", fig.align='left'}
# gene x spot
dim(dot.sample$srt$counts)

plot_data <- dot.sample$srt$coordinates
ggplot(plot_data, aes(x = Col, y = Row))+
  geom_point(size = 5)+
  theme_bw()+
  theme(panel.background = element_rect(fill = 'white'), 
        panel.grid = element_blank(),
        axis.text = element_blank(), 
        axis.title = element_blank(), 
        axis.ticks = element_blank())
```


# Decomposition

The first step of `DOT` is to set up a DOT object. `DOT` takes two main inputs:
  
  - `ref_data` is either a `matrix`, `dgMatrix`, or `data.frame` or a `Seurat`/`anndata` object containing the gene x cell count matrix for the reference single-cell data.
- If `ref_data` is a `matrix`, `dgMatrix`, or `data.frame`, then `ref_annotations` must be supplied as a 1-dimensional vector of cell annotations. 
- If `ref_data` is a `Seurat` or an `anndata` object, then `ref_annotations` can be either a 1-dimensional vector of cell annotations, a `character` indicating the `ident`/`obs` contained therein, or can be `NULL` in which case annotations are extracted from the object.

- `srt_data` is likewise either a `matrix`, `dgMatrix`, or `data.frame` or a `Seurat`/`anndata` object containing the gene x spot/cell count matrix for the target spatial data.
- If `srt_data` is a `matrix`, `dgMatrix`, or `data.frame`, then `srt_coords` must be supplied as a 2 x spot matrix of locations. 
- If `srt_data` is a `Seurat` or an `anndata` object, then `srt_coords` can be either 2 x spot matrix of locations or can be `NULL` in which case locations are extracted from the object.

In the following example, we supply the count matrices as `dgMatrix` objects and supply the annotations/locations explicitly.

```{r}
dot.srt <- setup.srt(srt_data = dot.sample$srt$counts, srt_coords = dot.sample$srt$coordinates)
dot.ref <- setup.ref(ref_data = dot.sample$ref$counts, ref_annotations = dot.sample$ref$labels, 10)

dot <- create.DOT(dot.srt, dot.ref)
```

We are now ready to perform decomposition:
```{r}
dot <- run.DOT.lowresolution(dot,  # The DOT object created above
                             lambda_a = 0,       # Abundance weight; a larger value more closely matches the abundance of cell types in the spatial data to those in the reference data
                             max_spot_size = 20, # Maximum size of spots (20 is usually sufficiently large for Visium slides) 
                             verbose = FALSE)

dim(dot@weights)
```

# Plotting

The output of DOT is contained in `dot@weights` which has spots as rows and cell types as columns, with each row denoting the cell type composition of the respective spot. A simple way of illustrating the cell type map is to annotate the spots based on the label of the most likely cell type.

Cell type map at the given annotation level:
  
```{r map_celltype, fig.height=5, fig.width=10.5, out.width="100%", fig.align='left'}
plot_data <- dot.sample$srt$coordinates
plot_data$celltype <- colnames(dot@weights)[apply(dot@weights, 1, which.max)]
ggplot(plot_data, aes(x = Col, y = Row, color = celltype))+
  geom_point(size = 5)+
  theme_bw()+
  theme(panel.background = element_rect(fill = 'white'), 
        panel.grid = element_blank(),
        axis.text = element_blank(), 
        axis.title = element_blank(), 
        axis.ticks = element_blank())
```

Cell type map at a higher annotation level:
  
```{r map_group, fig.height=5, fig.width=6.5, out.width="80%", fig.align='left'}
groups <- sapply(colnames(dot@weights), function(x) stringr::str_split(x, " ")[[1]][1])
agg_weights <- t(rowsum(t(dot@weights), groups))
plot_data$group <- colnames(agg_weights)[apply(agg_weights, 1, which.max)]

ggplot(plot_data, aes(x = Col, y = Row, color = group))+
  geom_point(size = 5)+
  theme_bw()+
  theme(panel.background = element_rect(fill = 'white'), 
        panel.grid = element_blank(),
        axis.text = element_blank(), 
        axis.title = element_blank(), 
        axis.ticks = element_blank())
```
